\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-04-03 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QLogic: fix commutative vs distributive

In this section we present our main argument, which is that the state space of quantum mechanics and the space of classical probability distributions possess a similar structure that allows one to apply commutative propositional calculus (i.e. ``classical logic'') on the first and non-commutative propositional calculus (i.e. ``quantum logic'') on the second. That is, both theories will have two lattices of propositions: a commutative one that follows classical logic and a non-commutative one that follows quantum logic.

Note that the classical space we will be using is not phase-space itself, but the space of probability distributions over phase-space, which is a real Hilbert space and thus has a similar structure to the quantum counterpart. The central point is that many features of quantum logic stem not from quantum mechanics per-se, but from the fact that quantum mechanics is a probabilistic theory, which means a classical probabilistic theory will share the same features. We find that the main point of confusion stems from comparing quantum theory, which is inherently a probabilistic theory, to a non-probabilistic classical theory. This approach does not allow one to understand which differences are due to the classical/quantum nature of the theory, and which are due simply to the probabilistic/non-probabilistic nature. By comparing quantum theory to a probabilistic classical theory, we put ourselves in a position to fully understand which difference is due to what, and have a much clearer understanding of both.\footnote{Note that probabilistic theories are important even in classical mechanics. Due to the nature of experimental verification, even in classical mechanics, there are always going to be uncertainties in the preparation and measurements, and repeated measurements will always lead to statistical constructs.} Critically, we will see that the non-commutativity of the quantum logic lattice does not originate from non-commuting observables or superposition of quantum mechanics: it is a property of the lattice of subspaces of any vector space and, indeed, of the lattice of sugroups of any group more in general. [CITE: Introduction to Lattices and Order Theory, Davey, Priestley - 2002 CUP]

Mathematically, the commutative lattice will be provided in both classical and quantum mechanics by the Borel algebra, the smallest $\sigma$-algebra that contains the topology; the non-commutative lattice will be provided by the set of subgroups ordered by inclusion. We will see that the non-commutative lattice is a strict subset of the commutative one, meaning that everything that can be described in quantum logic can also be described in classical logic. Moreover, the commutative lattice will include propositions, like propositions on the expectation value of observables, that are of scientific interest.\footnote{In an experimental context, in fact, by ``measurement'' one typically refers to the statistical collection of many ``single takes''. Quantum logic describes the single-take measurements. For example, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves.}

@@

To really understand the source of the non-commutativity, let us study the simplest possible example. Let us assume that there are only two possible classical states, $1$ and $2$. A state is identified by a two-valued distribution $[\rho_1, \rho_2]$, where $\rho_1$ is the probability to find the object in state 1 and $\rho_2$ is the probability to find the object in state 2. We note $[k,0]$ the subspace spanned by the first state only. This corresponds to the statement ``the system is certainly in state $1$'' since the probability for state $2$ is zero. Conversely, $[0,k]$ corresponds to the statement ``the system is certainly in state $2$''. The disjunction $[k,0] \vee [0,k]$ will correspond to the subspace spanned by both, which is the whole space $[k,j]$ where we can assign different non-zero probability to either case. We can also take the subspace $[k,k]$, the subspace spanned by vectors that have equal components in $1$ and $2$, which corresponds to the statement ``the system is equally likely to be in either state $1$ or $2$''. Note that $[k,0]$, $[0,k]$ and $[k,k]$ are all pairwise disjoint: no pairs have a vector in common apart from zero. Therefore their conjunctions $[k,0] \wedge [0,k] = [0,k] \wedge [k,k] = [k,k] \wedge [k,0] = [0,0]$ are zero dimensional. On the other hand, $[k,k]$ is a subspace of the whole space $[k,j]$ and therefore $[k,k] \wedge [k,j] = [k,k]$. Putting it all together, we have $[k,k] \wedge ( [k,0] \vee [0,k] ) = [k,k] \wedge [k,j] = [k,k]$. But we also have $( [k,k] \wedge [k,0] ) \vee ( [k,k] \wedge [0,k] ) = [0,0] \vee [0,0] = [0,0]$. Therefore the lattice is not distributive.

Note that during the discussion there was no mention as to whether $k$ and $j$ where real numbers, complex numbers or quaternions. The argument is formally the same in all cases. Therefore the lack of distributivity has nothing to do with the complexity of the Hilbert space or non-commuting observables. It's a property of the lattice of subspaces of any vector space and, more generally, of the lattice of subgroups of any group. In fact, consider the group of boost transformations in relativity. The boosts along one direction form a subgroup. If we take the subgroups of boost along $x$ and $y$ respectively, the smallest subgroup that contain both, the disjunction, is the subgroup of all boosts along any direction in the $x$-$y$ plane. As before, the disjunction is the span, so the resulting set is different from the union, and the union of all boosts along the x-axis and the y-axis is not a subgroup and therefore not contained in the lattice. Thus, this has the same non-commutative structure as quantum logic.

>>>>

In this section we present our main argument, which is that the state space of quantum mechanics and the space of classical probability distributions possess a similar structure that allows one to apply distributive propositional calculus (i.e. ``classical logic'') on the first and non-distributive propositional calculus (i.e. ``quantum logic'') on the second. That is, both theories will have two lattices of propositions: a distributive one that follows classical logic and a non-distributive one that follows quantum logic.

Note that the classical space we will be using is not phase-space itself, but the space of probability distributions over phase-space, which is a real Hilbert space and thus has a similar structure to the quantum counterpart. The central point is that many features of quantum logic stem not from quantum mechanics per-se, but from the fact that quantum mechanics is a probabilistic theory, which means a classical probabilistic theory will share the same features. We find that the main point of confusion stems from comparing quantum theory, which is inherently a probabilistic theory, to a non-probabilistic classical theory. This approach does not allow one to understand which differences are due to the classical/quantum nature of the theory, and which are due simply to the probabilistic/non-probabilistic nature. By comparing quantum theory to a probabilistic classical theory, we put ourselves in a position to fully understand which difference is due to what, and have a much clearer understanding of both.\footnote{Note that probabilistic theories are important even in classical mechanics. Due to the nature of experimental verification, even in classical mechanics, there are always going to be uncertainties in the preparation and measurements, and repeated measurements will always lead to statistical constructs.} Critically, we will see that the non-distributivity of the quantum logic lattice does not originate from non-commuting observables or superposition of quantum mechanics: it is a property of the lattice of subspaces of any vector space and, indeed, of the lattice of sugroups of any group more in general. [CITE: Introduction to Lattices and Order Theory, Davey, Priestley - 2002 CUP]

Mathematically, the distributive lattice will be provided in both classical and quantum mechanics by the Borel algebra, the smallest $\sigma$-algebra that contains the topology; the non-distributive lattice will be provided by the set of subgroups ordered by inclusion. We will see that the non-distributive lattice is a strict subset of the distributive one, meaning that everything that can be described in quantum logic can also be described in classical logic. Moreover, the distributive lattice will include propositions, like propositions on the expectation value of observables, that are of scientific interest.\footnote{In an experimental context, in fact, by ``measurement'' one typically refers to the statistical collection of many ``single takes''. Quantum logic describes the single-take measurements. For example, measurements of cross sections, the most typical type of measurement in the context of particle physics, are not part of the quantum lattice, and neither are measurements of the probabilities themselves.}

@@

To really understand the source of the non-distributivity, let us study the simplest possible example. Let us assume that there are only two possible classical states, $1$ and $2$. A state is identified by a two-valued distribution $[\rho_1, \rho_2]$, where $\rho_1$ is the probability to find the object in state 1 and $\rho_2$ is the probability to find the object in state 2. We note $[k,0]$ the subspace spanned by the first state only. This corresponds to the statement ``the system is certainly in state $1$'' since the probability for state $2$ is zero. Conversely, $[0,k]$ corresponds to the statement ``the system is certainly in state $2$''. The disjunction $[k,0] \vee [0,k]$ will correspond to the subspace spanned by both, which is the whole space $[k,j]$ where we can assign different non-zero probability to either case. We can also take the subspace $[k,k]$, the subspace spanned by vectors that have equal components in $1$ and $2$, which corresponds to the statement ``the system is equally likely to be in either state $1$ or $2$''. Note that $[k,0]$, $[0,k]$ and $[k,k]$ are all pairwise disjoint: no pairs have a vector in common apart from zero. Therefore their conjunctions $[k,0] \wedge [0,k] = [0,k] \wedge [k,k] = [k,k] \wedge [k,0] = [0,0]$ are zero dimensional. On the other hand, $[k,k]$ is a subspace of the whole space $[k,j]$ and therefore $[k,k] \wedge [k,j] = [k,k]$. Putting it all together, we have $[k,k] \wedge ( [k,0] \vee [0,k] ) = [k,k] \wedge [k,j] = [k,k]$. But we also have $( [k,k] \wedge [k,0] ) \vee ( [k,k] \wedge [0,k] ) = [0,0] \vee [0,0] = [0,0]$. Therefore the lattice is not distributive.

Note that during the discussion there was no mention as to whether $k$ and $j$ where real numbers, complex numbers or quaternions. The argument is formally the same in all cases. Therefore the lack of distributivity has nothing to do with the complexity of the Hilbert space or non-commuting observables. It's a property of the lattice of subspaces of any vector space and, more generally, of the lattice of subgroups of any group. In fact, consider the group of boost transformations in relativity. The boosts along one direction form a subgroup. If we take the subgroups of boost along $x$ and $y$ respectively, the smallest subgroup that contain both, the disjunction, is the subgroup of all boosts along any direction in the $x$-$y$ plane. As before, the disjunction is the span, so the resulting set is different from the union, and the union of all boosts along the x-axis and the y-axis is not a subgroup and therefore not contained in the lattice. Thus, this has the same non-distributive structure as quantum logic.
